---
title: "cyclistic bike share"
author: "Seth Gabienu"
date: "`r Sys.Date()`"
output: rmarkdown::github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message= FALSE)
```


## *Introduction*
this is a Google Data Analytics course project assignment organised by Coursera and Google. the assignment or case study sought to explore growth strategies of a fictitious Bike sharing company based in Chicago named Cyclistic bike. the company is an app-based bike sharing firm and has its market segmented into annual membership riders and casual riders. the marketing team has been tasked to identify strategies that would influence casual riders; riders who purchase a day ride pass or single ride pass, to annual membership.
as an analyst part of the marketing analytic team, we've been tasked by our marketing director to identify how bike usage between members and casual riders varies and recommend base on our analysis how the casual riders to be converted to annual members.

## *Background*
our exploration is into understanding how Cyclistic, a fictitious app based bike sharing company based in Chicago, approaches its marketing strategies and how recommendations can be made to enhance success of the organisation by increasing memberships segment of the firm.
Cyclistic bike share, founded in 2016, as at 2020 had a little over 5800 bikes, stationed in 692 locations in Chicago. the company has depended on the concept of going green by creating awareness and appealing to the general consumer segments. 
the firm has focused on flexible pricing plans by segmenting the market into; 
single-ride passes, full-day passes, annual memberships. the marketing team in simplifying the segments, grouped them into;
Cyclistic members for annual memberships segment,
casual riders for both riders that purchase single-ride pass and full day pass. 

the company has over the years of operation found out increasing their annual memberships will go a long way to expand the financial fortunes and has decided through the marketing department to identify strategies and ways of inducing casual riders to becoming annual members.
they also want to know from our analysis, how digital media could also help their marketing tactics.
As part of the course requirements, we will be using the Data Analytics phases of data analysis. they include;
ask, prepare, process, analyze, share, and act.


## *Ask*
Going by the data analytic phases in our lessons, we would first want to ask some questions of interest related to the business task. we would want to define the problem challenging our stakeholders (Cyclistic marketing analytic team, marketing director, the executive team). this will help make us focus on the project and avoid irrelevant issues.
in order to proceed to exploring our case study, we need to first understand the situation of the organisation by asking some pertinent questions pertaining to the bike usage offered by Cyclistic. this helps define the problem at hand.
finding the project outcome will require as to ask the following question, a way to be sure of our understanding of our case.
1. how are bikes usage between members and casual riders different?
this question would help us understand the significance of the worthiness of the case study.
2. what will motivate casual riders to become members?
3. how can Cyclistic use digital media to induce casual riders to becoming members?
although all questions are matters of interest to our project, emphasis will be mainly on how members and casual riders use bikes differently.

## *prepare*
This phase of our data analysis is concerned with the type of data required to resolve our questions and how to organise it to make it relevant for our business task. our dataset is a fictitious public data, made available by Motivate International Inc and we downloaded it from this link <https://divvy-tripdata.s3.amazonaws.com/index.html>
The link contains data on Cyclistic bike users from 2020 to 2023 but our interest according to our case assignment is to focus on a 12 month period. in doing this, we decide to focus on a 12 month calender year which is January to Decenber 2022.
Because of data privacy issues we are unable to see riders previous purchases to their credit card information for easy tracking. an examination of our data collected gives us a structured and organised data where data elements are related to one another and is also quantitative and wide in type.
Since our dataset contains the entire usage of bike for year 2022, data integrity relating to sampling bias or observer bias do not exist. since our data is coming from Motivate International Inc, we can verify the reliability, originality comprehensiveness, current state, and can also be cited. Also users were anonymized with ID numbers to obscure their identity. The data is stored in a comma delimiter format (.csv). Because we will be working in R studio, we will first load our packages.


### load packages
```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(tidyr)
library(lubridate)
library(geosphere)
```


### importing datasets from local

```{r}
library(readr)
jan<- read_csv("~/GDA/case study 1/2022/202201-divvy-tripdata.csv")
```
```{r}
library(readr)
feb<- read_csv("~/GDA/case study 1/2022/202202-divvy-tripdata.csv")
```
```{r}
library(readr)
mar<- read_csv("~/GDA/case study 1/2022/202203-divvy-tripdata.csv")
```
```{r}
library(readr)
april<- read_csv("~/GDA/case study 1/2022/202204-divvy-tripdata.csv")
```
```{r}
library(readr)
may<- read_csv("~/GDA/case study 1/2022/202205-divvy-tripdata.csv")
```
```{r}
library(readr)
june<- read_csv("~/GDA/case study 1/2022/202206-divvy-tripdata.csv")
```
```{r}
library(readr)
july<- read_csv("~/GDA/case study 1/2022/202207-divvy-tripdata.csv")
```
```{r}
library(readr)
aug<- read_csv("~/GDA/case study 1/2022/202208-divvy-tripdata.csv")
```
```{r}
library(readr)
sept<- read_csv("~/GDA/case study 1/2022/202209-divvy-publictripdata.csv")
```
```{r}
library(readr)
oct<- read_csv("~/GDA/case study 1/2022/202210-divvy-tripdata.csv")
```
```{r}
library(readr)
nov<- read_csv("~/GDA/case study 1/2022/202211-divvy-tripdata.csv")
```
```{r}
library(readr)
dec<- read_csv("~/GDA/case study 1/2022/202212-divvy-tripdata.csv")
```



## *Process*
in this phase of our data analytics, we would want to examine the integrity of our and tools we will be using for our analysis. It is worthwhile noting that after we downloaded our dataset, we replicated the dataset in an Excel format for future referencing. In this phase we would want to highlight tools we would be using to manipulate and clean our dataset. The following are tools employed for the studies.
1.	Microsoft Excel 
2.	R Studio
3.	GitHub

The process phase involves making the dataset ready through cleaning before it is transformed and analyzed. Since our data is stored in monthly format, we would have to merge all the months for a comprehensive analysis.

### combining all the data imported into a unified dataset
```{r}
trip_22<-bind_rows(jan, feb, mar, april, may, june, july, aug, sept, oct, nov, dec)
```


### cleaning the dataset
We checked for outliers by using “str(trip_22)” for entire dataset. Using the structure function provides a detailed informantion on the dataset. We noticed incomplete  entries appearing as “NA” .
We  therefore had to filter the incomplete entries and we observed 1,298,426 entries that had missing values in either a column or more than one columns.We cleaned the dataset  by dropping  incomplete rows as captured in the photo above.
The order of the date format in our original dataset (“started_at” and “ended_at”) is arranged as d/m/y_h:m and we need to format it to in order to be recognized by R in this format y/m/d_h:m

### inspecting column names
After merging all the monthly datasets our dataset tallied 5, 667,717 rows. When we inspected the column names we noticed the following;
"ride_id", "rideable_type", "started_at", "ended_at", "start_station_name", "start_station_id", "end_station_name", "end_station_id", "start_lat", "start_lng", "end_lat", "end_lng", "member_casual", “ride_length” and "day_of_week" (Total of 15  columns).

```{r}
head(trip_22)
```


### checking for outliers
```{r}
str(trip_22)
```


### checking for distinct output
```{r}
table(trip_22$rideable_type)  #total distinct values for ride types
table(trip_22$day_of_week)    # days of the week bikes used
table(trip_22$member_casual)  #unique values for mebers and casual riders
```


### dropping incomplete rows (NA) from our dataset
```{r}
clean_trip_22<-subset(trip_22, complete.cases(trip_22))  #clean dataset
```



### formatting started_at and ended_at variables
```{r}
clean_trip_22<-clean_trip_22 %>% mutate(started_at= format(dmy_hm(started_at)))%>% 
  mutate(ended_at= format(dmy_hm(ended_at)))  ##format started_at and ended_at date format 
```



### extract date, day of week, and month from started_at.
```{r}
clean_trip_22$date_started<-strftime(clean_trip_22$started_at, format= "%y-%m-%d") ##extract date

clean_trip_22$month<-strftime(clean_trip_22$started_at, format= "%B")  ## extract month

clean_trip_22$day_of_week<-strftime(clean_trip_22$date_started, format= "%A")  ##extract day of the week

```


### adding unique ride distance in meters
```{r}
clean_trip_22$ride_distance <- distGeo(matrix(c(clean_trip_22$start_lng, clean_trip_22$start_lat), ncol = 2), 
                                       matrix(c(clean_trip_22$end_lng, clean_trip_22$end_lat), ncol = 2)) ## ride distance in meters

```


### round ride_distance to 2 decimal points.
```{r}
clean_trip_22$ride_distance<-round(clean_trip_22$ride_distance, digits= 2)
```



### further cleaning: droping trip distance that is < 0
```{r}
clean_trip_22 <- clean_trip_22[!(clean_trip_22$ride_distance <= 0),]
```


## *Analyze*
This phase of the project is more into the transformation of the dataset to answer our assignment we have been tasked with. our project assignment is to explore how annual membership subscription riders use bike differently from casual riders. in this part, we would manipulate the dataset dscriptively to understand the variance and what can be done to achieve the objective of inducing casual riders to becoming annual members.


### a quick summary of our dataframe
```{r}
summary(clean_trip_22)
```



### examining the structure of the dataframe

```{r}
str(clean_trip_22)

```



### examining the average ride_length for the dataset
```{r}
clean_trip_22 %>%  
  summarise(avg_ride_length = mean(ride_length), median_length = median(ride_length), 
            max_ride_length = max(ride_length), min_ride_length = min(ride_length)) 
```



### checking the unique average ride length for both members and casual riders exclusively
```{r}
clean_trip_22 %>% group_by(member_casual) %>% 
  summarise(average_ride_length = mean(ride_length), median_length = median(ride_length), 
            max_ride_length = max(ride_length), min_ride_length = min(ride_length))

```


### Average ride length by months
```{r}
clean_trip_22 %>% group_by(member_casual, month) %>% 
  summarise(average_month_ride_length = mean(ride_length), median_month_length = median(ride_length), 
            max_month_ride_length = max(ride_length), min_month_ride_length = min(ride_length),.groups = "drop")
```


### average daily ride length
```{r}
clean_trip_22 %>% group_by(member_casual, day_of_week) %>% 
  summarise(average_day_ride_length = mean(ride_length), median_day_length = median(ride_length),max_day_ride_length = max(ride_length), min_day_ride_length=min(ride_length),.groups ="drop") %>% 
  arrange(member_casual, day_of_week)
```


### analyzing the average distance for the dataset
```{r}
clean_trip_22 %>% group_by(member_casual) %>% 
  summarise(average_ride_distance = mean(ride_distance), median_distance = median(ride_distance), 
            max_ride_distance = max(ride_distance), min_ride_distance = min(ride_distance))
```

### average monthly distance covered by members and casual riders
```{r}
clean_trip_22 %>% group_by(member_casual, month) %>% 
  summarise(average_month_ride_distance = mean(ride_distance), median_month_distance = median(ride_distance),max_month_ride_distance = max(ride_distance), min_month_ride_distance=min(ride_distance),.groups ="drop") %>% 
  arrange(member_casual, month)
```


### average daily ride distance
```{r}
clean_trip_22 %>% group_by(member_casual, day_of_week) %>% 
  summarise(average_day_ride_distance = mean(ride_distance), median_day_distance = median(ride_distance),max_day_ride_distance = max(ride_distance), min_day_ride_distance=min(ride_distance),.groups ="drop") %>% 
  arrange(member_casual, day_of_week)
```

### counting type of bike used 
```{r}
clean_trip_22 %>% 
  group_by(member_casual, rideable_type) %>% 
  summarise(number_rides=n(),.groups = "drop")
```
as can be seen, classic bikes were more used by members compared to casual riders. members did not use docked bikes, which is unlike casual riders



### checking count of members and casual bike riders.

```{r}
clean_trip_22%>% 
  group_by(member_casual) %>% 
  summarise(ride_count = length(ride_id), ride_percentage = (length(ride_id) / nrow(clean_trip_22)) * 100) 

```



## *Share*
this phase is concerned with presenting a graphical analysis to stakeholders involved in the project. 


### plottting count of bike users
we want to check the number of bike usage between members and casual riders. the graphics below shows that members patronised the bikes more than casual riders 
```{r}
ggplot(data = clean_trip_22)+
  geom_bar(mapping = aes(x=member_casual,fill=member_casual))+
  labs(title = "members vs casual bike users", caption = "https://divvy-tripdata.s3.amazonaws.com/index.html", x= "member_casual", y="ride count")

```

it can be seen from the chart that annual members tend to use bikes 20% more than casual riders



### plotting monthly ride distribution
this plot shows a breakdown into months of number of bike usage between members and casual riders.

```{r}
clean_trip_22 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_rides =n(),  average_ride_month=mean(ride_length),.groups = "drop") %>% 
  arrange(month, member_casual) %>% 
  ggplot(aes(x=month, y=number_rides, fill=member_casual))+
  labs(title = "monthly usage: members vs casúals")+
  geom_col(width = 0.6, position = position_dodge(width = 0.4))+
  scale_y_continuous(labels = function(x) format(x, scientific=FALSE))
```

an examination of the monthly distribution of bike usage shows that annual members usage of bike was very high during June, July, August, which can partly be explained by the summer period. we also noticed similarly, casual riders also recorded high usage during the same period but unlike annual members.



### plotting daily usage of bikes by members and casual riders

```{r}
clean_trip_22 %>% 
  group_by(member_casual, day_of_week) %>%  #groups by member_casual
  summarise(number_of_rides = n() #calculates the number of rides and average duration 
            ,average_ride_length = mean(ride_length),.groups="drop") %>% 
  arrange(day_of_week,member_casual) %>%
    ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  labs(title ="Daily ride used per day") +
  geom_col(width=0.5, position = position_dodge(width=0.3)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

the daily usage presenting an interesting dynamics. we observed that, casual riders tend to use the bike more than annual members on Saturdays.



### plotting type of bikes used
we want to check the different types of bike used by members and casual riders
```{r}
clean_trip_22 %>% 
  group_by(member_casual, rideable_type) %>% 
  summarise(number_rides =n(),.groups = "drop") %>% 
  arrange(rideable_type, member_casual) %>% 
  ggplot(aes(x=rideable_type, y=number_rides, fill=member_casual))+
  labs(title = "types of bikes: members vs casúals")+
  geom_col(width = 0.6, position = position_dodge(width = 0.4))+
  scale_y_continuous(labels = function(x) format(x, scientific=FALSE))
```

we discovered that while both riders used both classic bikes and electric bikes, casual riders also used docked bikes.



### Average ride length per day in seconds
```{r}
clean_trip_22 %>%  
  group_by(member_casual, day_of_week) %>% 
  summarise(average_ride_length = mean(ride_length), .groups="drop") %>%
  ggplot(aes(x = day_of_week, y = average_ride_length, fill = member_casual)) +
  geom_col(width=0.5, position = position_dodge(width=0.3)) + 
  labs(title ="Average ride length in a day")
```

we noticed that casual riders throughout the week spent more time on average biking than annual members, with weekends being their busiest periods



### plotting average time for bike usage in a month

```{r}
clean_trip_22 %>%  
  group_by(member_casual, month) %>% 
  summarise(average_ride_length = mean(ride_length), .groups="drop") %>%
  ggplot(aes(x = month, y = average_ride_length, fill = member_casual)) +
  geom_col(width=0.4, position = position_dodge(width=0.4)) + 
  labs(title ="Average ride time in a month in seconds")
```

on monthly, casual riders spent more time riding especially in March, April, and May 2022, compared to annual members who mainly spent about half usage time of casual riders




## *Act*
so far, our analysis' captured in the plots gives an interesting picture.
first,  we discovered that the average annual usage time of the bikes for causal riders was about twice the usage time for annual members with 1348 seconds and 741 seconds respectively.
 this shows that annual members spent less time riding unlike casual riders.
 on monthly basis, casual riders spent a lot time riding in March, April, and May and weekend tend to be periods casual riders spent a lot of time biking.
Also, it is evident that the average annual distance covered by casual riders was 2.2% higher than annual members, with both averaging 2324.723 meters and 2132.218 meters respectively. 
However, bike usage between both casual riders and annual members showed a 39.21% and 60.79% accordingly. 
this explains that annual members usage of bikes was more than 20% higher than casual riders. 
we observed that bike usage was high during the summer periods for both riders but the distribution was higher for annual members through the month.
we found out that casual riders had a high demand for bikes on weekend unlike annual members with spike in demand during the weekdays.
Finaly, it emerged that annual members had high preference for classic bikes with no interest in docked bikes.

While the data provides mixed and unique findings, we observed that the uniqueness of ride id to a ride makes it difficult to track a casual rider's usage of bike for a period. as a result would have help explore the frequency an individual rider uses the bike on weekly or monthly basis. this we believe would have help narrow the casual riders to target.
Thus, having knowledge about how each casual rider uses bike and providing weekly or monthly reports via the app about bike usage would segment and target casual riders who have high demand for the service. 
we also suggest monthly subscriptions for users who may not need the service for the entire year due to weather changes and other factors but may demand the bikes for about a week or more.
